### 2017-05-08 2017mergingmerges/section/05-Evaluation.tex
consistently use Figure N instead of Fig. N

The first group in Fig.~\ref{fig:bench:all} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
The hand-fused version is the fastest, while our version is slower, but still faster than the vector version.
The fact that our version is slower than the hand-fused version is surprising, as the generated GHC core is almost identical: the only difference is an extra continuation bound inside the loop, which acts as a sort of indirect jump in the loop body.
% It is possible that recent work on optimising continuations in GHC~\cite{downen2016sequent} will solve this.

The first file benchmark simply appends two files, while counting the lines.
We compared against three Haskell streaming libraries: Conduit, Pipes, and Streaming.
The second group in Fig.~\ref{fig:bench:all} shows the runtimes for appending 2MB of data.
The absolute performance here is poor because all are using line-buffered IO; in practice one would use chunked IO, but the overhead per chunk would remain.

The second file benchmark takes a file and partitions it into two files: one with even-length lines, and one with odd-length lines.
The output lines are also counted.
The three streaming libraries are pull-based, and do not support multiple outputs: the program must be written in a convoluted way or at least partially hand-fused.
Even with hand-fusion, the Pipes and Conduit programs are slower than ours, as well as losing any abstraction benefits from using a streaming library.
The final group in Fig.~\ref{fig:bench:all} shows the runtimes for partitioning a 1MB file.

% Fused result programs like the one in Fig. \ref{fig:Process:Fused} can be lowered directly to a target language, such as Haskell, C or LLVM abstract assembly. The details of such lowering surely affect the performance of the final program, but as they are largely unrelated to the fusion algorithm itself, we focus on the form of the fused program expressed directly in the process language.

@@

% -----------------------------------------------------------------------------
%\subsection{Result Size}
%
%As with any fusion system, we must be careful that the size of the result code does not become too large when more and more processes are fused together. The left of Fig.~\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner. To produce this graph we programmatically generated dataflow networks for \emph{all possible} pipelined combinations of the @map@, @filter@, @scan@, @group@ and @merge@ combinators, and tried all possible fusion orders consiting of adjacent pairs of processes. The @merge@ combinator itself has two inputs, so only works at the very start of the pipeline --- we present result for pipelines with and without a @merge@ at the start. The right of Fig.~\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream. In both cases the number of states in the result process grows linearly with the number of processes. In all combinations, with up to 7 processes there are less than 100 states in the result process. 
%%%% AR: Note that the split version also has only one merge?
%
%The size of the result process is roughly what one would get when inlining the definitions of each of the original source processes. This is common with other systems based on inlining and/or template meta-programming, and is not prohibitive.
%
%On the other hand, Fig.~\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number of input processes. The source dataflow networks consists of N merge processes, N+1 input streams, and a single output stream. The output of each merge process is the input of the next, forming a chain of merges. In source notation the network for N = 3 is @sOut = merge sIn1 (merge sIn2 (merge sIn3 sIn4))@.
%
%When fusing two processes the fusion algorithm essentially compares every state in the first process with every state in the second, computing a cross product. During the fusion transform, as states in the result process are generated they are added to a finite map --- the @instrs@ field of the process definition. The use of the finite map ensures that identical states are always combined, but genuinely different states always make it into the result. 
%
%In the worst case, fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each. If we assume the two processes have about the same number of states then this is O($n^2$). Fusing the next process into this result yields O($n^3$), so overall the worst case number of states in the result will be O($n^k$), where $k$ is the number of processes fused. 
%
%In the particular case of @merge@, the implementation has two occurrences of the @push@ instruction. During fusion, the states for the consuming process are inlined at each occurrence of @push@. These states are legitimately different because at each occurence of @push@ the input channels of the merge process are in different channel states, and these channel states are included in the overall process state.

% -----------------------------------------------------------------------------
\subsection{Optimisation and Drop Instructions}
\label{s:Optimisation}
After we have fused two processes together, it may be possible to simplify the result before fusing in a third. Consider the result of fusing @group@ and @merge@ which we saw back in Fig.~\ref{fig:Process:Fused}. At labels @F1@ and @F2@ are two consecutive @jump@ instructions. The update expressions attached to these instructions are also non-interfering, which means we can safely combine these instructions into a single @jump@. In general, we prefer to have @jump@ instructions from separate processes scheduled into consecutive groups, rather than spread out through the result code. The (PreferJump) clauses of Fig.~\ref{fig:Fusion:Def:StepPair} implement a heuristic that causes jump instructions to be scheduled before all others, so they tend to end up in these groups.

@@

% Another optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable. In Fig.~\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction. Remember from \S\ref{s:Fusion:FusingPulls} that the variable @b1@ is the stream buffer variable. Values pulled from stream @sIn1@ are first stored in @b1@ before being copied to @v@ and @x1@. When the two processes to be fused share a common input stream, use of stream buffer variable allows one process to continue using the value that was last pulled from the stream, while the other moves onto the next one.

When the two processes are able to accept the next variable from the stream at the same time, there is no need for the separate stream buffer variable. This is the case in Fig.~\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@. To increase the chance that we can perform copy-propagation, we need both processess to want to pull from the same stream at the same time. Moving the @drop@ instruction for a particular stream as late as possible prevents a @pull@ instruction from a second process being scheduled in too early.

>>>>

The first group in Figure~\ref{fig:bench:all} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
The hand-fused version is the fastest, while our version is slower, but still faster than the vector version.
The fact that our version is slower than the hand-fused version is surprising, as the generated GHC core is almost identical: the only difference is an extra continuation bound inside the loop, which acts as a sort of indirect jump in the loop body.
% It is possible that recent work on optimising continuations in GHC~\cite{downen2016sequent} will solve this.

The first file benchmark simply appends two files, while counting the lines.
We compared against three Haskell streaming libraries: Conduit, Pipes, and Streaming.
The second group in Figure~\ref{fig:bench:all} shows the runtimes for appending 2MB of data.
The absolute performance here is poor because all are using line-buffered IO; in practice one would use chunked IO, but the overhead per chunk would remain.

The second file benchmark takes a file and partitions it into two files: one with even-length lines, and one with odd-length lines.
The output lines are also counted.
The three streaming libraries are pull-based, and do not support multiple outputs: the program must be written in a convoluted way or at least partially hand-fused.
Even with hand-fusion, the Pipes and Conduit programs are slower than ours, as well as losing any abstraction benefits from using a streaming library.
The final group in Figure~\ref{fig:bench:all} shows the runtimes for partitioning a 1MB file.

% Fused result programs like the one in Figure \ref{fig:Process:Fused} can be lowered directly to a target language, such as Haskell, C or LLVM abstract assembly. The details of such lowering surely affect the performance of the final program, but as they are largely unrelated to the fusion algorithm itself, we focus on the form of the fused program expressed directly in the process language.

@@

% -----------------------------------------------------------------------------
%\subsection{Result Size}
%
%As with any fusion system, we must be careful that the size of the result code does not become too large when more and more processes are fused together. The left of Figure~\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner. To produce this graph we programmatically generated dataflow networks for \emph{all possible} pipelined combinations of the @map@, @filter@, @scan@, @group@ and @merge@ combinators, and tried all possible fusion orders consiting of adjacent pairs of processes. The @merge@ combinator itself has two inputs, so only works at the very start of the pipeline --- we present result for pipelines with and without a @merge@ at the start. The right of Figure~\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream. In both cases the number of states in the result process grows linearly with the number of processes. In all combinations, with up to 7 processes there are less than 100 states in the result process. 
%%%% AR: Note that the split version also has only one merge?
%
%The size of the result process is roughly what one would get when inlining the definitions of each of the original source processes. This is common with other systems based on inlining and/or template meta-programming, and is not prohibitive.
%
%On the other hand, Figure~\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number of input processes. The source dataflow networks consists of N merge processes, N+1 input streams, and a single output stream. The output of each merge process is the input of the next, forming a chain of merges. In source notation the network for N = 3 is @sOut = merge sIn1 (merge sIn2 (merge sIn3 sIn4))@.
%
%When fusing two processes the fusion algorithm essentially compares every state in the first process with every state in the second, computing a cross product. During the fusion transform, as states in the result process are generated they are added to a finite map --- the @instrs@ field of the process definition. The use of the finite map ensures that identical states are always combined, but genuinely different states always make it into the result. 
%
%In the worst case, fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each. If we assume the two processes have about the same number of states then this is O($n^2$). Fusing the next process into this result yields O($n^3$), so overall the worst case number of states in the result will be O($n^k$), where $k$ is the number of processes fused. 
%
%In the particular case of @merge@, the implementation has two occurrences of the @push@ instruction. During fusion, the states for the consuming process are inlined at each occurrence of @push@. These states are legitimately different because at each occurence of @push@ the input channels of the merge process are in different channel states, and these channel states are included in the overall process state.

% -----------------------------------------------------------------------------
\subsection{Optimisation and Drop Instructions}
\label{s:Optimisation}
After we have fused two processes together, it may be possible to simplify the result before fusing in a third. Consider the result of fusing @group@ and @merge@ which we saw back in Figure~\ref{fig:Process:Fused}. At labels @F1@ and @F2@ are two consecutive @jump@ instructions. The update expressions attached to these instructions are also non-interfering, which means we can safely combine these instructions into a single @jump@. In general, we prefer to have @jump@ instructions from separate processes scheduled into consecutive groups, rather than spread out through the result code. The (PreferJump) clauses of Figure~\ref{fig:Fusion:Def:StepPair} implement a heuristic that causes jump instructions to be scheduled before all others, so they tend to end up in these groups.

@@

% Another optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable. In Figure~\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction. Remember from \S\ref{s:Fusion:FusingPulls} that the variable @b1@ is the stream buffer variable. Values pulled from stream @sIn1@ are first stored in @b1@ before being copied to @v@ and @x1@. When the two processes to be fused share a common input stream, use of stream buffer variable allows one process to continue using the value that was last pulled from the stream, while the other moves onto the next one.

When the two processes are able to accept the next variable from the stream at the same time, there is no need for the separate stream buffer variable. This is the case in Figure~\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@. To increase the chance that we can perform copy-propagation, we need both processess to want to pull from the same stream at the same time. Moving the @drop@ instruction for a particular stream as late as possible prevents a @pull@ instruction from a second process being scheduled in too early.
